Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2025-05-25 manuscript/manuscript.tex
Edit abstract

\textbf{Conclusion:} All models performed well at the population level, while GPT-3.5/4 and InSilicoVA performed well for some CODs. GPT models are unable to replace physician coding yet, but have made large improvements over InterVA-5, setting the foundation for assisting physicians with alternative diagnoses to reduce ill-defined codes and disagreement when coding VAs.
}

@@

Another research opportunity is in the integration of GPT, InterVA-5, and InSilicoVA models into VA systems to assist physicians in COD assignment. In dual-coded VA systems (described in Section \ref{methods-data}), two physicians are randomly assigned to each record and require second inspections of each other's assignment (reconciliation) and evaluation by a third more senior physician if their assignments do not agree. As mentioned in Section \ref{discuss-adv-hscav}, suggestion of alternative assignments from GPT and InSilicoVA models potentially reduces the disagreement between physicians, and ill-defined records, while allowing physicians to focus on more difficult records. Thus, model suggestions can be integrated into VA systems by presenting COD suggestions to physicians after their initial COD assignment, which allows them to consider alternative assignments and possibly revise their assignments based on the suggestions. At step 2 in Figure \ref{fig-discuss-phyassist}, GPT, InterVA-5, and InSilicoVA models can suggest COD assignments to consider, providing the option in step 2b to revise or proceed with their initial assignment. Our future work will be a first step in computer assisted verbal autopsy, assessing the effects of these model suggestions on improve VA data quality (e.g. increase in agreed records, reduction of ill-defined deaths). In preparation, we have integrated GPT-4, InterVA-5, and InSilicoVA model suggestions into our HEAL-SL study after survey round 2 \citep{njalauniversityHealthySierraLeone2023}.

>>>>

\textbf{Conclusion:} All models performed well at the population level, while GPT-3.5/4 and InSilicoVA performed well for some CODs. GPT models have yet to replace physician coding, but have made improvements over InSilicoVA and InterVA-5. Thus, our research lays the foundation for future opportunities in computer assisted VA, where physicians are presented with alternative COD assignments from computer models to help reduce ill-defined codes and physician disagreement.
}

@@

Another research opportunity is in the integration of GPT, InterVA-5, and InSilicoVA models into VA systems to assist physicians in COD assignment. In dual-coded VA systems (described in Section \ref{methods-data}), two physicians are randomly assigned to each record and require second inspections of each other's assignment (reconciliation) and evaluation by a third more senior physician if their assignments do not agree. As mentioned in Section \ref{discuss-adv-hscav}, suggestion of alternative assignments from GPT and InSilicoVA models potentially reduces the disagreement between physicians, and ill-defined records, while allowing physicians to focus on more difficult records. Thus, model suggestions can be integrated into VA systems by presenting COD suggestions to physicians after their initial COD assignment, which allows them to consider alternative assignments and possibly revise their assignments based on the suggestions. At step 2 in Figure \ref{fig-discuss-phyassist}, GPT, InterVA-5, and InSilicoVA models can suggest COD assignments to consider, providing the option in step 2b to revise or proceed with their initial assignment. Our future work will be a first step in computer assisted verbal autopsy, assessing the effects of these model suggestions on improve VA data quality (e.g. increase in agreed records, reduction of ill-defined deaths). In preparation, we have integrated GPT-4, InterVA-5, and InSilicoVA model suggestions into our on-going HEAL-SL study after survey round 2 \citep{njalauniversityHealthySierraLeone2023} with goals of increasing physician agreement and reducing ill-defined COD assignments.
